---
title: "part2"
author: "Hans Elliott"
date: "4/13/2022"
output: github_document
---
# Chapter 11: It's Learning
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, I turn the spiral data generation into a callable function so we can easily mess with the scale of the data. (Idea courtesy of NNFS of course)
```{r sim-data}
## Creating Data
# Source:
sim_data_fn = function(
  N = 100, # number of points per class
  D = 2, # dimensionality (number of features)
  K = 3, # number of classes
  random_order = FALSE
){
   X = data.frame() # data matrix (each row = single sample)
   y = data.frame() # class labels

for (j in (1:K)){
  r = seq(0.05,1,length.out = N) # radius
  t = seq((j-1)*4.7,j*4.7, length.out = N) + rnorm(N, sd = 0.3) # theta
  Xtemp = data.frame(x =r*sin(t) , y = r*cos(t)) 
  ytemp = data.frame(matrix(j, N, 1))
  X = rbind(X, Xtemp)
  y = rbind(y, ytemp)
  }
spiral_data = cbind(X,y)
colnames(spiral_data) = c(colnames(X), 'label')

# Want randomly ordered labels?
if (random_order==TRUE) {spiral_data$label = sample(1:3, size = N*K, 
                                                    replace = TRUE)}
return(spiral_data)
}
```

```{r start-nn-fns}
### Initialize Parameters for a Layer
init_params = function(n_inputs = "# of features", n_neurons){
        
    weights = matrix(data = (0.1 * rnorm(n = n_inputs*n_neurons)),
                             nrow = n_inputs, ncol = n_neurons)
           #Number of weights = the number of inputs*number of neurons,
           #since every connection between the previous neurons (from input) and
           #the current neurons have an associated weight
    biases = matrix(data = 0, nrow = 1, ncol = n_neurons)
           #Number of biases = the number
    #saves:
    list("weights"=weights, "biases"=biases)
}


### Dense Layers ----
layer_dense = list(
## FORWARD PASS FUNCTION 
 forward = function(    #default vals for weights/biases are random
            inputs, 
            n_neurons, 
            parameters
            #weights = 
            #    matrix(data = (0.10 * rnorm(n = ncol(inputs)*n_neurons)),
            #                 nrow = ncol(inputs), ncol = n_neurons), 
            #biases = 
            #    matrix(data = 0, nrow = 1, ncol = n_neurons)
 ) {#begin fn
   
  n_inputs = ncol(inputs)
      #determine number of inputs per sample from dims of the input matrix
      #should be equal to # of features (i.e, columns) in a sample (i.e, a row)
  
  #Initalize Weights and Biases
  #if (exists("new_weights")){
  #  weights = new_weights
  #} else {
  #  weights = matrix(data = (0.10 * rnorm(n = n_inputs*n_neurons)),
  #                   nrow = n_inputs, ncol = n_neurons)
  #}
       #Number of weights = the number of inputs*number of neurons. 
       #(Random numbers multipled by 0.10 to keep small.)
  #if (exists("new_biases")){
  #  biases = new_biases
  #} else {
  #  biases = matrix(data = 0, nrow = 1, ncol = n_neurons)
  #}
       #bias will have shape 1 by number of neurons. we initialize with zeros
 weights = parameters$weights
 biases = parameters$biases
 #Forward Pass 
 output = inputs%*%weights + biases[col(inputs%*%weights)]
 
 #UPDATE: 
 #need to save the inputs for backprop, so we create a list of objects.
 
 list("output" = output, #for forward pass
      "inputs" = inputs, "weights"= weights, "biases" = biases) #for backprop
 #and prints output by default, but invisibly
 #invisible(output)

 },
## BACKWARD PASS
 backward = function(inputs, weights, dvalues){
    #Gradients on parameters
    dweights = t(inputs)%*%dvalues
    dbiases = colSums(dvalues)
    #Gradient on values
    dinputs = dvalues%*%t(weights) 
    #save:
    list("dinputs"=dinputs,
         "dweights"=dweights,
         "dbiases"=dbiases)
 }
 
)
### Activation Functions ----
## ReLU
activation_ReLU = list(
  #FORWARD PASS
  forward = function(input_layer){
    
    output = matrix(sapply(X = input_layer, 
                    function(X){max(c(0, X))}
                    ), 
                  nrow = nrow(input_layer), ncol = ncol(input_layer))
    #ReLU function coerced into a matrix so the shape
    #is maintained (it will be equivalent to that of the input shape)
    
    #Function saves:
    list("output" = output, "inputs" = input_layer)
    #And prints
    #invisible(output)
  },
  #BACKWARD PASS
  backward = function(inputs, dvalues){
    dinputs = dvalues
    dinputs[inputs <= 0] = 0
    #save:
    list("dinputs"=dinputs)
  }
)

## SoftMax
activation_Softmax = list(
  forward = function(inputs){
          #scale inputs
          max_value = apply(X = inputs, MARGIN = 2,  FUN = max)
          scaled_inputs = sapply(X = 1:ncol(inputs), 
                 FUN = function(X){
                    inputs[,X] - abs(max_value[X])})

          # exponetiate
          exp_values = exp(scaled_inputs)
          # normalize
          norm_base = matrix(rowSums(exp_values),
                             nrow = nrow(inputs), ncol = 1)
          probabilities = sapply(X = 1:nrow(inputs),
                          FUN = function(X){exp_values[X,]/norm_base[X,]}) 
          return(t(probabilities))
          #(transpose probabilities)
          },
  
  backward = function(softmax_output, dvalues){
    #*INCOMPLETE SECTION - don't use*
    #flatten output array
    flat_output = as.vector(softmax_output)
    
    #calculate jacobian matrix of output
    jacobian_matrix = diag(flat_output) - flat_output%*%t(flat_output)
    
    #calculate sample-wise gradient
    dinputs = jacobian_matrix%*%flat_dvalues
    
    
  }
)

### Loss ----
Categorical_CrossEntropy = list(
  #FORWARD PASS
  forward = function(y_pred = "softmax output", y_true = "targets"){
    
    #DETECT NUMBER OF SAMPLES
    samples = length(y_true)  

    #CLIP SAMPLES TO AVOID -Inf ERROR
    y_pred_clipped = ifelse(y_pred <= 1e-7, 1e-7, 
                        ifelse(y_pred >= (1-1e-7), (1-1e-7), y_pred))
    

    #DETERMINE IF Y_TRUE IS ONE-HOT-ENCODED AND SELECT CORRESPODNING CONFIDENCES
    confidences = ifelse(nrow(t(y_true)) == 1, 
      #if y_true is a single vector of labels (i.e, sparse), then confidences =
                    y_pred_clipped[cbind(1:samples, y_true)],
                      
                      ifelse(nrow(y_true) > 1,
                      #else, if y_true is one-hot encoded, then confidences =
                             rowSums(y_pred_clipped*y_true),
                             #else
                             "error indexing the predicted class confidences")
                  )
                    
    #CALC LOSS FOR EACH SAMPLE (ROW)
    neg_log_likelihoods = -log(confidences)
    return(neg_log_likelihoods)
    
  },
  #BACKWARD PASS
  backward = function(y_true, dvalues){
    #number of samples
    samples = length(dvalues)
    
    #number of labels
    labels = length(unique(dvalues[1,]))
    
    #if labels are sparse, turn them into one-hot encoded vector
    y_true = ifelse(#if
                    nrow(t(y_true)) ==1,
                    #one-hot-encode
                    y_true = do.call(rbind,
                                     lapply(X = y_true,
                                       function(X) as.integer(
                                          !is.na(match(unique(
                                          unlist(y_true)
                                                        ), X)
                                                ))
                                      )), 
                    #else
                    y_true)
    
    #calculate gradient
    dinputs = -y_true/dvalues
    #normalize gradient
    dinputs = dinputs/samples
    return(dinputs)
  }
)
#---
#Softmax X Cross Entropy- combined softmax activation
# & cross-entropy loss for faster backprop
activation_loss_SoftmaxCrossEntropy = list(
  #FORWARD PASS
  forward = function(inputs, y_true){
    #output layer's activation function
    softmax_out = activation_Softmax$forward(inputs)
    #calculate loss
    loss = Categorical_CrossEntropy$forward(softmax_out, y_true)
    #function saves:
    list("softmax_output"=softmax_out, "loss"=loss) 
  },
  #BACKWARD PASS
  backward = function(dvalues, y_true){
    
    #Detect number of samples
    if (is.vector(dvalues)) {      #if one sample
      samples = 1
    } else if (is.array(dvalues)) {  #else if multiple samples
      samples = nrow(dvalues)
    } else print("error checking shape of inputs")
    
    #Reverse One-Hot Encoding
    #if labels are one-hot encoded, turn them discrete values
     ##helper function 
    anti_ohe = function(y_true){
               unique_classes = ncol(y_true)
               samples = nrow(y_true)
               y_true_vec = as.vector(y_true)
                    
               class_key = rep(1:unique_classes, each = samples)
               y_true = class_key[y_true_vec==1]
                    #selects the classes that correspond to 1s in y_true vector
                    return(y_true)
                    }
     ##check & modify
    y_true = if(is.array(y_true)){ #if one-hot encoded
                    #change to sparse
                    anti_ohe(y_true)
              } else y_true
    
    #Calculate gradient
     #Copy so we can modify
     dinputs = dvalues
     #Calculate gradient
     #index the prediction array with the sample number and its
     #true value index, subtracting 1 from these values. Requires discrete,
     #not one-hot encoded, true labels (explaining the need for the above step)
     dinputs[cbind(1:samples, y_true)] = dinputs[cbind(1:samples, y_true)] - 1
     #Normalize gradient
     dinputs = dinputs/samples
    #save desired outputs
    list("dinputs" = dinputs, "samples" = samples, "y_true" = y_true)
  }
)

```


```{r SGD-optimizer-basic}
optimizer_SGD = list(
  update_params = function(layer_forward, layer_backward, 
                           learning_rate = 1){ #default value
    #current params
    current_weights = layer_forward$weights
    current_biases = layer_forward$biases
    #gradients
    weight_gradients = layer_backward$dweights
    bias_gradients = layer_backward$dbiases
    #update
    weights = current_weights - learning_rate*weight_gradients
    biases = current_biases - learning_rate*bias_gradients
      #save:
      list("weights" = weights, "biases" = biases)
  }
)
```


First, note that I've added a parameter initialization stage that occurs outside of the loop. The resulting list of parameter objects is then passed into each corresponding layer.   
```
init_params = function(n_inputs = "# of features", n_neurons){
        
    weights = matrix(data = (0.1 * rnorm(n = n_inputs*n_neurons)),
                             nrow = n_inputs, ncol = n_neurons)
           #Number of weights = the number of inputs*number of neurons,
           #since every connection between the previous neurons (from input) and
           #the current neurons have an associated weight
    biases = matrix(data = 0, nrow = 1, ncol = n_neurons)
           #Number of biases = the number
    #saves:
    list("weights"=weights, "biases"=biases)
}

```
**EPOCH LOOP**  
Note: I use a learning rate of 1 here, which is equivalent to no learning rate.
```{r epoch-loop-1}
set.seed(1) ##set seed for results replication

#Inputs
spiral_data = sim_data_fn(N=33) #33 obs per class, total of 99 obs.
spiral_X = spiral_data[,c("x","y")]
spiral_X = as.matrix(spiral_X)

#Initalize Weights & Biases Outside of Loop
l1_params = init_params(n_inputs = 2,      ## ncol(spiral_X) would also work
                         n_neurons = 64)   ## = to desired # neurons in layer 
l2_params = init_params(n_inputs = 64,     ## = to n_neurons in prior layer
                         n_neurons = 3)    ## = to desired # neurons in layer

# EPOCH LOOP
tot_epochs = 1500L

for (epoch in 1:tot_epochs) {
#forward
layer1 = layer_dense$forward(inputs = spiral_X, 
                             n_neurons = 64,
                             parameters = l1_params)
                                    #weights = l1_params$weights,
                                     # biases = l1_params$biases)
  layer1_relu = activation_ReLU$forward(input_layer = layer1$output)
      layer2 = layer_dense$forward(inputs = layer1_relu$output, 
                                   n_neurons = 3,
                                   parameters = l2_params)
          output = 
          activation_loss_SoftmaxCrossEntropy$forward(inputs = layer2$output,
                                                  y_true = spiral_data$label)
              #metrics:
              LOSS = output$loss
              PRED = max.col(output$softmax_output, ties.method = "random") 
              ACC = mean(PRED==spiral_data$label)

          #backward
          loss_softm_b = activation_loss_SoftmaxCrossEntropy$backward(
                                      dvalues = output$softmax_output, 
                                      y_true = spiral_data$label)
      l2_b = layer_dense$backward(inputs = layer2$inputs,
                            weights = layer2$weights,
                            dvalues = loss_softm_b$dinputs)
  l1_relu_b = activation_ReLU$backward(inputs = layer1_relu$inputs,
                                       dvalues = l2_b$dinputs)     
l1_b = layer_dense$backward(inputs = layer1$inputs,
                            weights = layer1$weights,
                            dvalues = l1_relu_b$dinputs)      

#optimize
l1_optim_params = optimizer_SGD$update_params(layer_forward = layer1,
                                       layer_backward = l1_b,
                                       learning_rate = 1)
                                       
l2_optim_params = optimizer_SGD$update_params(layer_forward = layer2,
                                       layer_backward = l2_b,
                                       learning_rate = 1)
#update weights & biases
l1_params = l1_optim_params
l2_params = l2_optim_params

#repeat w new weights & biases
##  
#Status Report:
  if (epoch == 1){print(c("epoch","ovrflw err","loss","accuracy"))}
  if (epoch %in% seq(0,tot_epochs,by=100)){
        print(c( epoch, 
                 sum(is.nan(output$softmax_output)),#make sure NaN error is gone
                 LOSS,
                 ACC) )}


#Save Final Metrics:
  if (epoch==tot_epochs){
             out = list("loss"=LOSS,
                         "accuracy"=ACC,
                         "predictions"=PRED)}
}#end loop
```


```{r}
out$loss
out$accuracy
out$predictions
```
Wow! The neural net was able to almost perfectly learn the labels of the spiral data in under 1000 epochs. Of course this wasn't a very difficult dataset for the model to learn, but we've come a long way.  
A note:
In a previous attempt, I trained the model on a dataset of 300 obs. and with a learning rate of 1, the NN broke down after about 1000 epochs. Specifically, it became filled with NaN values, rendering it useless. I'm guessing that this is related to exploding gradients. I stopped the NN around epoch 900 and found that the size of the weights had grown to be quite large. By about epoch 1000, it would start returning NaN values. This can be solved using a lower learning rate (e.g, 0.5 worked). In practice, a learning rate of 1 is quite high, allowing every step in the gradient descent process to be rather large, directly leading to gradient explosion (NNFS, Ch. 10 pg 24). 
This problem is relatively common and can be solved in [various ways](https://machinelearningmastery.com/exploding-gradients-in-neural-networks/). My initial tendency is to re-code something so that this simply cannot happen, which points towards _gradient clipping_ or _weight regularization_, but alternative options (lower learning rates or reshuffling the shape of the network) are also viable solutions, so I won't touch anything for now.  
Either way, this is a good segue into adding some further hyperparameter options to the Neural Network.   


```{r}
#Actual
plot(spiral_data$x, spiral_data$y, col = spiral_data$label)
#Predicted
plot(spiral_data$x, spiral_data$y, col = out$predictions)
```


# Chapter 12: Hyperparameters
## Learning Rate Decay
Learning rate decay describes a process whereby the learning rate decreases during the course of model training. 
This could be achieved by programmatically or manually adjusting the learning rate when the model's loss plateaus or begins jumping randomly.  
It could also be achieved by programming a simple **Decay Rate**, which lowers the learning rate each epoch or each batch. This is also known as 1/t decaying or exponential decaying. For example:  

```{r decay-ec}
start_learn_rate = 1
learn_rate_decay = 0.1
for (step in 1:20){
  learning_rate = start_learn_rate * (1 / (1 + learn_rate_decay*step))
                                      ## ensures that learn rate never gets >1
  print(learning_rate)
}
```

We will simply add this as an optional hyperparameter to our SGD optimizer. Since I'm currently going with the for loop method, iteration will just be set equal to the current epoch.  
The default values ensure that if we decided to ignore decay, it will not affect the learning rate.  
```{r sgd-with-decay}
optimizer_SGD = list(
  update_params = function(layer_forward, layer_backward, 
                           learning_rate = 1,
                           decay = 0, iteration = 1){ #default values
    #current params
    current_weights = layer_forward$weights
    current_biases = layer_forward$biases
    #gradients
    weight_gradients = layer_backward$dweights
    bias_gradients = layer_backward$dbiases
    
    #learning rate
    currnt_learn_rate = learning_rate * (1 / (1 + decay*iteration)) 
    
    #update
    weights = current_weights - currnt_learn_rate*weight_gradients
    biases = current_biases - currnt_learn_rate*bias_gradients
      #save:
      list("weights" = weights, "biases" = biases, "lr" = currnt_learn_rate)
  }
)
```

Here I implement our training loop using a decay rate of 0.001, starting with a learning rate of 1. I train the model to learn a dataset of 300 observations, which previously did not work with the high learning rate due to exploding gradients.  
Adding the decay rate solved this exploding gradients issue, and the model trains successfully for 1500 epochs.  
```{r loop-w-decay}
set.seed(1) ##set seed for results replication

#Inputs
spiral_data = sim_data_fn(N=100) #100 obs per class, total of 300 samples
spiral_X = spiral_data[,c("x","y")] #leave out label
spiral_X = as.matrix(spiral_X)

#Initalize Weights & Biases Outside of Loop
l1_params = init_params(n_inputs = 2,      ## ncol(spiral_X) would also work
                         n_neurons = 64)   ## = to desired # neurons in layer 
l2_params = init_params(n_inputs = 64,     ## = to n_neurons in prior layer
                         n_neurons = 3)    ## = to desired # neurons in layer

# EPOCH LOOP
tot_epochs = 1500L

for (epoch in 1:tot_epochs) {
#forward
layer1 = layer_dense$forward(inputs = spiral_X, 
                             n_neurons = 64,
                             parameters = l1_params)
  layer1_relu = activation_ReLU$forward(input_layer = layer1$output)
      layer2 = layer_dense$forward(inputs = layer1_relu$output, 
                                   n_neurons = 3,
                                   parameters = l2_params)
          output = 
          activation_loss_SoftmaxCrossEntropy$forward(inputs = layer2$output,
                                                  y_true = spiral_data$label)
              #metrics:
              LOSS = output$loss
              PRED = max.col(output$softmax_output, ties.method = "random") 
              ACC = mean(PRED==spiral_data$label)

          #backward
          loss_softm_b = activation_loss_SoftmaxCrossEntropy$backward(
                                      dvalues = output$softmax_output, 
                                      y_true = spiral_data$label)
      l2_b = layer_dense$backward(inputs = layer2$inputs,
                            weights = layer2$weights,
                            dvalues = loss_softm_b$dinputs)
  l1_relu_b = activation_ReLU$backward(inputs = layer1_relu$inputs,
                                       dvalues = l2_b$dinputs)     
l1_b = layer_dense$backward(inputs = layer1$inputs,
                            weights = layer1$weights,
                            dvalues = l1_relu_b$dinputs)      

#optimize
l1_optim_params = optimizer_SGD$update_params(layer_forward = layer1,
                                       layer_backward = l1_b,
                                       learning_rate = 1,
                                       decay = .001,iteration = epoch
                                       )
l2_optim_params = optimizer_SGD$update_params(layer_forward = layer2,
                                       layer_backward = l2_b,
                                       learning_rate = 1,
                                       decay = .001, iteration = epoch)
#update weights & biases
l1_params = l1_optim_params
l2_params = l2_optim_params

#repeat w new weights & biases
##  
#Status Report:
  if (epoch == 1){print(c("epoch","ovrflw err","loss","accuracy", "learn rate"))}
  if (epoch %in% seq(0,tot_epochs,by=100)){
        print(c( epoch, 
                 sum(is.nan(output$softmax_output)),#make sure NaN error is gone
                 LOSS,
                 ACC,
                 l1_optim_params$lr) )} #(include learn rate)


#Save Final Metrics:
  if (epoch==tot_epochs){
             out = list("loss"=LOSS,
                         "accuracy"=ACC,
                         "predictions"=PRED)}
}#end loop
```
The learning rate decreases quite quickly, but this seems to work. Rather than losing control, the NN steadily decreases the loss and reaches an accuracy of 98%, which is much better than random guessing. It seems to get stuck in a local minimum by the 1000th epoch.  

## Momentum
Momentum allows the size and direction of each "downward" step in gradient descent to depend upon the size and direction of prior steps. This can help the model pass through local loss minimums, making it more likely that a global minimum is found. I find this easiest to understand [visually](https://nnfs.io/jog/). You can see how the change in the loss function "speeds up" when loss is decreasing rapidly, and slows down when loss starts increasing.  
To actually implement this method, we will allow some fraction of the previous parameter updates (i.e, the changes to weights & biases) to be combined with a fraction of the current parameter updates. Thus, the portion of the gradient from preceding steps - the direction of change decided in previous epochs - determines the actual momentum.  

Now, we can add this as a hyperparameter. Due to how I am executing my loop and updating parameters within the loop, I've decided to implement momentum as an option in the paramter initialization function. If set to `TRUE`, the function will generate a matrix of zeros of the right shape. Also notice that I've added a cache option using the same methodology. This will be explained further later.  
```{r revamped-initparams}
init_params = function(n_inputs = "# of features",
                       n_neurons,
                       momentum = FALSE,
                       cache = FALSE){
        
    weights = matrix(data = (0.1 * rnorm(n = n_inputs*n_neurons)),
                             nrow = n_inputs, ncol = n_neurons)
           #Number of weights = the number of inputs*number of neurons,
           #since every connection between the previous neurons (from input) and
           #the current neurons have an associated weight
    biases = matrix(data = 0, nrow = 1, ncol = n_neurons)
           #Number of biases = the number
    
    #momentum initialization
    weight_momentums = matrix(data = 0, 
                            nrow = nrow(weights),
                            ncol = ncol(weights))
    bias_momentums = matrix(data = 0,
                            nrow = nrow(biases),
                            ncol = ncol(biases))
    #cache initialization
    weight_cache = matrix(data = 0,
                          nrow = nrow(weights),
                          ncol = ncol(weights))
    bias_cache = matrix(data = 0,
                        nrow = nrow(biases),
                        ncol = ncol(biases))

    #saving:
    if (momentum == TRUE & cache == FALSE){ ##momentums only
          list("weights"=weights,"biases"=biases,
               "weight_momentums"=weight_momentums,
               "bias_momentums"=bias_momentums)
    } else if (cache == TRUE & momentum == FALSE){ ##cache only
          list("weights"=weights,"biases"=biases,
               "weight_cache"=weight_cache,
               "bias_cache"=bias_cache)
    } else if (momentum == TRUE &  cache == TRUE){ ##momentums and cache
          list("weights"=weights,"biases"=biases,
               "weight_momentums"=weight_momentums,
               "bias_momentums"=bias_momentums,
               "weight_cache"=weight_cache,
               "bias_cache"=bias_cache)
    } else if (momentum == FALSE & cache == FALSE){ ##no momentums or cache
          list("weights"=weights,"biases"=biases)
    }
}

```

Now we can update the SGD optimizer to utilize momentum. If the momentum rate is set above 0 (should be between 0 and 1), then the optimizer will update weights and biases using the momentum method. This will work _only if_ momentum was enabled in the parameter initialization function (above).  
If momentum was enabled but the rate stays at zero, the optimizer will still calculate the updates using the momentum method but this will be equivalent to the "vanilla" method.
```{r revamped-SGD}
optimizer_SGD = list(
  update_params = function(layer_forward, layer_backward, #required
                           learning_rate = 1,
                           decay = 0, iteration = 1,
                           momentum_rate = 0 
                           ){
    #current params
    current_weights = layer_forward$weights
    current_biases = layer_forward$biases
    #gradients
    weight_gradients = layer_backward$dweights
    bias_gradients = layer_backward$dbiases
    
    #learning rate
    currnt_learn_rate = learning_rate * (1 / (1 + decay*iteration)) 
    
    #param updates with momentum
     #If momentum == TRUE in parameter initialization, then weights_momentum
     #(and implictly, bias_momentum) will exist
    if (exists("weight_momentums", where = layer_forward)) {
      #current momentums
      weight_momentums = layer_forward$weight_momentums
      bias_momentums = layer_forward$bias_momentums
      
      #Update weights & biases with momentum:
      #Take prior updates X retainment factor (the "momentum rate"),
      #and update with current gradients
      weight_update = 
        (momentum_rate*weight_momentums) - (currnt_learn_rate*weight_gradients)
      bias_update = 
        (momentum_rate*bias_momentums) - (currnt_learn_rate*bias_gradients)
      #update params with the calculated updates
      weights = current_weights + weight_update
      biases = current_biases + bias_update
      #also update momentums
      weight_momentums = weight_update
      bias_momentums = bias_update
      #save:
      list("weights" = weights, "biases" = biases, "lr" = currnt_learn_rate,
           "weight_momentums" = weight_momentums, 
           "bias_momentums" = bias_momentums)
    } else {
      
    #param updates without momentum (vanilla)
      #calculate updates
      weight_update = -currnt_learn_rate*weight_gradients
      bias_update = -currnt_learn_rate*bias_gradients
      #apply updates
      weights = current_weights + weight_update
      biases = current_biases + bias_update
      #save:
      list("weights" = weights, "biases" = biases, "lr" = currnt_learn_rate)
    }
    
}
)

```

Finally, in order for this to all work seamlessly we need to update the layer_dense function to store `weight_momentums` and `bias_momentums` when momentum is enabled. Above and below you can see that I utilized the base::exists() function. In these cases, it is telling R to look for an object called "weights_momentums" within the parameters object. If it exists (which it will if momentum is enabled in the param initialization function) then each layer will store the matrices of momentums.
```{r revamped-denselayer}
### Dense Layers ----
layer_dense = list(
## FORWARD PASS FUNCTION 
 forward = function(    #default vals for weights/biases are random
            inputs, 
            n_neurons, 
            parameters
 ) {#begin fn
   
 n_inputs = ncol(inputs)
 weights = parameters$weights
 biases = parameters$biases
 #Forward Pass 
 output = inputs%*%weights + biases[col(inputs%*%weights)]
 
 #If momentum == TRUE & cache == FALSE in parameter initialization, 
 #then layer saves momentum only
 if (exists(x = "weight_momentums", where = parameters) & 
     !exists(x = "weight_cache", where = parameters)){
 list("output" = output, ##for forward pass
      "inputs" = inputs, "weights"= weights, "biases" = biases, ##for backprop
      "weight_momentums"=parameters$weight_momentums, ##for momentum
      "bias_momentums"=parameters$bias_momentums)
 #if momentum==FALSE & cache==TRUE, saves cache only
 } else if (!exists(x = "weight_momentums", where = parameters) &
            exists(x = "weight_cache", where = parameters)){
 list("output" = output, ##for forward pass
      "inputs" = inputs, "weights"= weights, "biases" = biases, ##for backprop
      "weight_cache"=parameters$weight_cache, ##for cache
      "bias_cache"=parameters$bias_cache)
 #if momentum==TRUE & cache==TRUE, saves both
 } else if (exists(x = "weight_momentums", where = parameters) &
            exists(x = "weight_cache", where = parameters)){
 list("output" = output, ##for forward pass
      "inputs" = inputs, "weights"= weights, "biases" = biases, ##for backprop
      "weight_momentums"=parameters$weight_momentums, ##for momentum
      "bias_momentums"=parameters$bias_momentums,
      "weight_cache"=parameters$weight_cache, ##for cache
      "bias_cache"=parameters$bias_cache)
 #else both==FALSE, ignore momentum & cache
 } else {
   #otherwise, just save
 list("output" = output, ##for forward pass
      "inputs" = inputs, "weights"= weights, "biases" = biases) ##for backprop
 }

 },#end fwd

## BACKWARD PASS
 backward = function(inputs, weights, dvalues){
    #Gradients on parameters
    dweights = t(inputs)%*%dvalues
    dbiases = colSums(dvalues)
    #Gradient on values
    dinputs = dvalues%*%t(weights) 
    #save:
    list("dinputs"=dinputs,
         "dweights"=dweights,
         "dbiases"=dbiases)
 }
 
)#end list
```

Okay, to sum up what just happened here:  
If we want to use momentum, we set momentum==TRUE in the paramter intialization stage (and when we want to use cache, we will do the same). Then, when the params are passed into the layer objects, their will exists objects that correspond to matrices for the weights and bias momentums. Since these objects exist, the layer function will save them.  
Then, when the layer is passed into the optimizer in the "layer_forward" option, the optimizer will detect the existence of the momentum objects and use them to calculate updates with momentums. It will also use whatever momentum_rate the user sets.
Let's try it:

```{r loop-w-momentum}
set.seed(1) ##set seed for results replication

#Inputs
spiral_data = sim_data_fn(N=100) #100 obs per class, total of 300 samples
spiral_X = spiral_data[,c("x","y")] #leave out label
spiral_X = as.matrix(spiral_X)

#Initalize Weights & Biases Outside of Loop
l1_params = init_params(n_inputs = 2,     ## ncol(spiral_X) would also work
                        n_neurons = 64,   ## = to desired # neurons in layer 
                        momentum = TRUE)
l2_params = init_params(n_inputs = 64,    ## = to n_neurons in prior layer
                        n_neurons = 3,    ## = to desired # neurons in layer
                        momentum = TRUE)

# EPOCH LOOP
tot_epochs = 2000L

for (epoch in 1:tot_epochs) {
#forward
layer1 = layer_dense$forward(inputs = spiral_X, 
                             n_neurons = 64,
                             parameters = l1_params)
  layer1_relu = activation_ReLU$forward(input_layer = layer1$output)
      layer2 = layer_dense$forward(inputs = layer1_relu$output, 
                                   n_neurons = 3,
                                   parameters = l2_params)
          output = 
          activation_loss_SoftmaxCrossEntropy$forward(inputs = layer2$output,
                                                  y_true = spiral_data$label)
              #metrics:
              LOSS = output$loss
              PRED = max.col(output$softmax_output, ties.method = "random") 
              ACC = mean(PRED==spiral_data$label)

          #backward
          loss_softm_b = activation_loss_SoftmaxCrossEntropy$backward(
                                      dvalues = output$softmax_output, 
                                      y_true = spiral_data$label)
      l2_b = layer_dense$backward(inputs = layer2$inputs,
                            weights = layer2$weights,
                            dvalues = loss_softm_b$dinputs)
  l1_relu_b = activation_ReLU$backward(inputs = layer1_relu$inputs,
                                       dvalues = l2_b$dinputs)     
l1_b = layer_dense$backward(inputs = layer1$inputs,
                            weights = layer1$weights,
                            dvalues = l1_relu_b$dinputs)      

#optimize
l1_optim_params = optimizer_SGD$update_params(layer_forward = layer1,
                                       layer_backward = l1_b,
                                       learning_rate = 1,
                                       decay = .001,iteration = epoch,
                                       momentum_rate = 0.5)
l2_optim_params = optimizer_SGD$update_params(layer_forward = layer2,
                                       layer_backward = l2_b,
                                       learning_rate = 1,
                                       decay = .001, iteration = epoch,
                                       momentum_rate = 0.5)
#update weights & biases
l1_params = l1_optim_params
l2_params = l2_optim_params

  #repeat w new weights,biases, & momentums...
##  
#Status Report:
  if (epoch == 1){print(c("epoch","ovrflw err","loss","accuracy", "learn rate"))}
  if (epoch %in% seq(0,tot_epochs,by=100)){
        print(c( epoch, 
                 sum(is.nan(output$softmax_output)),#make sure NaN error is gone
                 LOSS,
                 ACC,
                 l1_optim_params$lr) )} #(include learn rate)


#Save Final Metrics:
  if (epoch==tot_epochs){
             out = list("loss"=LOSS,
                         "accuracy"=ACC,
                         "predictions"=PRED)}
}#end loop

```

Adding momentum gave us a slight boost. We managed to get out of the local minimum and this improved the accuracy. 0.9933 corresponds to 298/300 correctly classified samples.   

```{r}
#Actual
plot(spiral_data$x, spiral_data$y, col = spiral_data$label)
#Predicted
plot(spiral_data$x, spiral_data$y, col = out$predictions)

```

# The AdaGrad Optimizer
There are other optimization methods we can use during the learning process. One commonly known method is AdaGrad, or adaptive gradient, which is a variant of SGD that uses a "per-parameter" learning rate rather than a globally shared one.  By saving the history of all parameter updates
> "AdaGrad provides a way to normalize parameter updates by keeping a history of previous updates — the bigger the sum of the updates is, in either direction (positive or negative), the smaller updates are made further in training. This lets less-frequently updated parameters to keep-up with changes, effectively utilizing more neurons for training" (NNFS, Ch.10 p. 52)  

This is implemented by storing the (squared) parameter gradients in a cache object, and then adding the new (squared) parameter gradients to the cache during each iteration. From the cache, the actual parameter updates are calculated.
Pseudo-code:  
```
cache = cache + param_gradient^2
param_updates = learning_rate * param_gradient/(sqrt(cache) + epsilon)
```
By dividing the parameter gradient by the square root of the cache, we effectively normalize the updates made to the weights and biases. Squaring the parameter gradient in the first step ensures that the cache values are all positive, so that when we divide them from the parameter gradient value the signs do not switch. By taking the square root of the cache (after squaring the parameter gradients), we ensure that the cache values grow slower. Finally, epsilon is a hyperparameter that prevents us from dividing the gradient by zero. It is typically set to be quite small.
One note: as training evolves, the size of the cache values mechanically grow, making the parameter updates (i.e, learning) smaller and smaller. Therefore, AdaGrad isn't a widely used optimizer.
Still, we'll implement it to better learn how it works and so that we have the option to use it.

```{r adagrad-code}
#NOTE: must initialize params with cache==TRUE
optimizer_AdaGrad = list(
  update_params = function(layer_forward, layer_backward, #required
                           learning_rate = 1,
                           decay = 0, iteration = 1,
                           epsilon = 1e-7 
                           ){
    #current params
    current_weights = layer_forward$weights
    current_biases = layer_forward$biases
    #gradients
    weight_gradients = layer_backward$dweights
    bias_gradients = layer_backward$dbiases
    
    #learning rate
    currnt_learn_rate = learning_rate * (1 / (1 + decay*iteration)) 
    
    #cache
    ##update cache with squared current gradients
    weight_cache = layer_forward$weight_cache + weight_gradients^2
    bias_cache = layer_forward$bias_cache + bias_gradients^2
    
    #SGD param updates with normalization 
      #calculate updates
      weight_update = -currnt_learn_rate*weight_gradients /
                        (sqrt(weight_cache) + epsilon)
      bias_update = -currnt_learn_rate*bias_gradients /
                        (sqrt(bias_cache) + epsilon)
      #apply updates
      weights = current_weights + weight_update
      biases = current_biases + bias_update
        #save:
        list("weights" = weights, "biases" = biases, "lr" = currnt_learn_rate,
             "weight_cache" = weight_cache, "bias_cache" = bias_cache)
} ##these are the new params to be passed on to the layers when loop restarts
)
```


# The RMSProp Optimizer
Another adaptation of SGD is Root Mean Square Propagation, which sort of combines the ideas of momentum and adaptive, per-parameter learning rates. Each update to the saved parameter updates (the cache) retains a fraction of the current cache while being updated with a fraction of the new parameter gradients. Ideally, this prevents learning from stalling as it may with the AdaGrad optimizer, since we are not continually summing squared gradients. 
Pseudo-code:
```
cache = rho*cache + (1 - rho)*(param_gradient^2)
```
Rho, another hyperparameter, can be thought of as the cache memory decay rate. If 0, the optimizer will utilize none of the previous updates, and all of the current gradients to calculate parameter updates. If 1, then none of the current updates would be used (and hence learning would not work). Therefore, we will set rho to be somewhere between 0 and 1.  
According to NNFS, the learning rate should be set fairly low when using this optimizer, since it will already have a lot of "momentum" from the cache-stored updates when rho is high. They suggest a default learning rate of 0.001, which is the default value for RMSProp "in well-known machine learning frameworks" (pg. 58).  


```{r rmsprop-code}
#NOTE: must initialize params with cache==TRUE
optimizer_RMSProp = list(
  update_params = function(layer_forward, layer_backward, #required
                           learning_rate = 0.001,
                           decay = 0, iteration = 1,
                           epsilon = 1e-7,
                           rho = 0.9
                           ){
    #current params
    current_weights = layer_forward$weights
    current_biases = layer_forward$biases
    #gradients
    weight_gradients = layer_backward$dweights
    bias_gradients = layer_backward$dbiases
    
    #learning rate
    currnt_learn_rate = learning_rate * (1 / (1 + decay*iteration)) 
    
    #cache
    ##update cache with squared current gradients
    weight_cache = rho * layer_forward$weight_cache + 
                  (1-rho) * weight_gradients^2
    bias_cache = rho * layer_forward$bias_cache + 
                  (1-rho) * bias_gradients^2
    
    #SGD param updates with normalization 
      #calculate updates
      weight_update = -currnt_learn_rate*weight_gradients /
                        (sqrt(weight_cache) + epsilon)
      bias_update = -currnt_learn_rate*bias_gradients /
                        (sqrt(bias_cache) + epsilon)
      #apply updates
      weights = current_weights + weight_update
      biases = current_biases + bias_update
        #save:
        list("weights" = weights, "biases" = biases, "lr" = currnt_learn_rate,
             "weight_cache" = weight_cache, "bias_cache" = bias_cache)
} ##these are the new params to be passed on to layers when the loop restarts
)
```




# The Adam Optimizer
Adam, or Adaptive Momentum, is a widely-used optimizer. When I was first getting into deep learning and trying to figure out what optimizer to use, most resources pointed to Adam, so it's exciting that we get to implement such a commonly used method.    
Adam builds on RMSProp and directly incorporates momentums into its optimization calculations. Instead of applying the current gradients as the parameter updates, we're going to apply momentums, and then apply a per-parameter adaptive learning rate with the cache, as in RMSProp.  
Adam also adds a "bias correction mechanism" which intends to compensate for initializing momentum and cache values as zeros. This is implemented by dividing both momentum and cache by $1 - beta^{step}$. Beta should be a number between 0 and 1, so that as step increases (i.e, as training progresses) $beta^{step}$ converges (asymptotes) to zero, and the bias correction term effectively disappears (becomes 1). Essentially, the initial momentums and cache will be divided by a number between 1 and 0 (a fraction), causing them to be larger, which is meant to speed up training in the early training iterations.  

```{r adam-code}
#NOTE: must intialize paramters with both cache AND momentum
optimizer_Adam = list(
  update_params = function(layer_forward, layer_backward, #required
                           learning_rate = 0.001,
                           decay = 0, iteration = 1,
                           epsilon = 1e-7,
                           beta_1 = 0.9, beta_2 = 0.999
                           ){
    #current params
    current_weights = layer_forward$weights
    current_biases = layer_forward$biases
    #gradients
    weight_gradients = layer_backward$dweights
    bias_gradients = layer_backward$dbiases
    
    #learning rate
    currnt_learn_rate = learning_rate * (1 / (1 + decay*iteration)) 
    
    #momentums
    #update momentums with current gradients and bias correct
    weight_momentums = 
      (beta_1*layer_forward$weight_momentums + (1-beta_1)*weight_gradients ) /
                              (1 - (beta_1^iteration )) ##bias correction
    
    bias_momentums = 
      (beta_1*layer_forward$bias_momentums + (1-beta_1)*bias_gradients) /
                              (1 - (beta_1^iteration))
                    

    #cache
    #update cache with squared gradients and bias correct
    weight_cache = 
      (beta_2*layer_forward$weight_cache + (1-beta_2)*weight_gradients^2) /
                              (1 - (beta_2^iteration)) ##bias correction

    bias_cache = 
      (beta_2*layer_forward$bias_cache + (1-beta_2)*bias_gradients^2) /
                              (1 - (beta_2^iteration))
    
    #calculate param updates (with momentums, and normalize with cache)
    weight_update = -currnt_learn_rate*weight_momentums /
                        (sqrt(weight_cache) + epsilon)
    bias_update = -currnt_learn_rate*bias_momentums /
                        (sqrt(bias_cache) + epsilon)
    
    #apply updates
    weights = current_weights + weight_update
    biases = current_biases + bias_update
      #save:
      list("weights" = weights, "biases" = biases, "lr" = currnt_learn_rate,
           "weight_momentums" = weight_momentums, 
           "bias_momentums" = bias_momentums,
           "weight_cache" = weight_cache,
           "bias_cache" = bias_cache)

}##these are the updated params to be passed back into the layers
)
```




## Testing the changes
Optimizers can be easily switched out in the optimization phase of the loop. Remember, for AdaGrad and RMSProp, just set cache==TRUE, since they require a cache of weights and biases. For Adam, set momentum==TRUE and cache==TRUE, since it uses both. (For vanilla SGD, set both to FALSE, the default).  

```{r epoch-loop-adam}
set.seed(1) ##set seed for results replication

#Inputs
spiral_data = sim_data_fn(N=100) #100 obs per class, total of 300 samples
spiral_X = spiral_data[,c("x","y")] #leave out label
spiral_X = as.matrix(spiral_X)

#Initalize Weights & Biases Outside of Loop
## Set momentum & cache to TRUE to use Adam optimizer
l1_params = init_params(n_inputs = 2,     ## ncol(spiral_X) would also work
                        n_neurons = 64,   ## = to desired # neurons in layer 
                        momentum = TRUE, cache = TRUE)
l2_params = init_params(n_inputs = 64,    ## = to n_neurons in prior layer
                        n_neurons = 3,    ## = to desired # neurons in layer
                        momentum = TRUE, cache = TRUE)

# EPOCH LOOP
tot_epochs = 2000L

for (epoch in 1:tot_epochs) {
#forward
layer1 = layer_dense$forward(inputs = spiral_X, 
                             n_neurons = 64,
                             parameters = l1_params)
  layer1_relu = activation_ReLU$forward(input_layer = layer1$output)
      layer2 = layer_dense$forward(inputs = layer1_relu$output, 
                                   n_neurons = 3,
                                   parameters = l2_params)
          output = 
          activation_loss_SoftmaxCrossEntropy$forward(inputs = layer2$output,
                                                  y_true = spiral_data$label)
              #metrics:
              LOSS = output$loss
              PRED = max.col(output$softmax_output, ties.method = "random") 
              ACC = mean(PRED==spiral_data$label, na.rm=T)

          #backward
          loss_softm_b = activation_loss_SoftmaxCrossEntropy$backward(
                                      dvalues = output$softmax_output, 
                                      y_true = spiral_data$label)
      l2_b = layer_dense$backward(inputs = layer2$inputs,
                            weights = layer2$weights,
                            dvalues = loss_softm_b$dinputs)
  l1_relu_b = activation_ReLU$backward(inputs = layer1_relu$inputs,
                                       dvalues = l2_b$dinputs)     
l1_b = layer_dense$backward(inputs = layer1$inputs,
                            weights = layer1$weights,
                            dvalues = l1_relu_b$dinputs)      

#optimize
l1_optim_params = optimizer_Adam$update_params(layer_forward = layer1,
                                       layer_backward = l1_b,
                                       learning_rate = 0.02,
                                       decay = 0,iteration = epoch,
                                       epsilon = 1e-7, 
                                       beta_1 =  0.5, beta_2 = 0.5)
l2_optim_params = optimizer_Adam$update_params(layer_forward = layer2,
                                       layer_backward = l2_b,
                                       learning_rate = 0.02,
                                       decay = 0, iteration = epoch,
                                       epsilon = 1e-7, 
                                       beta_1 = 0.5, beta_2 = 0.5)
#update weights & biases
l1_params = l1_optim_params
l2_params = l2_optim_params

  #repeat w new weights,biases, & momentums...
##  
#Status Report:
  if (epoch == 1){print(c("epoch","ovrflw err","loss","accuracy", "learn rate"))}
  if (epoch %in% seq(0,tot_epochs,by=100)){
        print(c( epoch, 
                 sum(is.nan(output$softmax_output)),#make sure NaN error is gone
                 LOSS,
                 ACC,
                 l1_optim_params$lr) )} #(include learn rate)


#Save Final Metrics:
  if (epoch==tot_epochs){
             out = list("loss"=LOSS,
                         "accuracy"=ACC,
                         "predictions"=PRED)}
}#end loop

```

The different optimizers can easily be tested by switching out the optimizer and setting the appropriate arguments in the `#optimize` stage of the loop. I tested each individually but chose just to show Adam. One thing to note is that if the learning rate is set too high, these optimizers are bound to run into exploding gradient issues which, in practical terms, means that values will start to turn to NaN ("not a number") and the neural net will break down. Beyond that, the optimizers works quite well, and Adam is extremely effective.  
Adam is the only optimizer thus far that managed to classify every single sample correctly in a 300 sample dataset (which is tricky, since the higher number of observations leads to more overlapping of classes).  


https://www.r-bloggers.com/2016/01/strategies-to-speedup-r-code/






